P3064R1: How to Avoid OOTA Without Really Trying

### 2024-07-12 ootavacuous.tex
Add advice per Michael Wong feedback from St. Louis
Explicitly and prominently state that nothing need change for the
common-case implementations in the scope of this paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

We begin with a brief overview of the OOTA problem, followed
by an equally brief summary of prior work in this area, and ending
with a quick overview of the \co{herd7} tool that will be used to
evaluate litmus tests.

@@

Because nothing else in the C++ memory model rules out such OOTA cycles,
the C++ standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, but those of us
writing code in the real world must rely on actual C++ implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations require checks on such
	speculation, and these checks must be based on actual values
	loaded.}

@@

>>>>

We begin with a summary of what needs to change and a brief overview of
the OOTA problem, followed by an equally brief summary of prior work in
this area, and ending with a quick overview of the \co{herd7} tool that
will be used to evaluate litmus tests.

\subsection{What Needs to Change?}
\label{sec:What Needs to Change?}

What changes are needed to prevent OOTA from C++ implementations using
traditional compilers and traditional hardware, including CPUs and
GPGPUs?

\begin{itemize}
\item	To the standard:
	No changes.
	The indefinite prohibition of out-of-thin-air values needs
	to remain.
\item	To implementations:
	No changes.
	Future implementations must avoid optimizations that invent,
	duplicate, or repurpose atomic loads.
	However, such optimizations must be avoided for other reasons,
	including preserving basic laws of arithmetic
	(see Appendix~\ref{app:Inventing Atomic Loads} for more detail).
\item	To user code:
	No changes.
	Users must avoid undefined behavior in portable code, just as
	is the case now.
\end{itemize}

In short, no changes are needed because current implementations already
avoid OOTA cycles.
The point of this paper is to give a semi-formal demonstration of
why this is the case.

@@

Because nothing else in the C++ memory model rules out such OOTA cycles,
the C++ standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, for example, to
prevent misapplication of the as-if rule in oracular C++ implementations,
but those of us writing code in the real world must rely on actual C++
implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations require checks on such
	speculation, and these checks must be based on actual values
	loaded.}

@@

This work shows that avoiding OOTA cycles requires no changes to the
standard, to current implementions, or to user practices for portable
code.
Future implementations must avoid optimizations that invent, duplicate,
or repurpose atomic loads.
More work is required to delineate any further restrictions on future
implementations that might carry out optimizations based on global
inter-thread analysis.
